Fast Sorting on Modern Computers

### 2005-08-31 docs/tech_report/conclusions.tex
david's final changes

Among them is the fact that bimodal branch predictors have performed better than
2-level adaptive ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by 2-level adaptive predictors - do
not exist, a bimodal predictors perform better.

@@

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a four
heap lowered our instruction count. Our level 2 misses also concurred with
LaMarca, with only one difference: memory-tuned heapsort, being an out-of-place sort,
fills the cache fully with a smaller number of keys than base heapsort, which is
in-place. LaMarca's results do not show this. Nor do they show a difference in
cache misses between these two sorts when the array does fit in the cache.
LaMarca's base heapsort have the characteristics of an out-of-place heapsort.
However, our results show that his implicit heap improvements have led to a
significantly faster heapsort.

@@

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $H_N-1$ branch mispredictions per key, which applies
to long arrays, though short arrays become more predictable the longer they
get, as in heapsort; tiling mergesort can be done in two steps, reducing the
cache miss count; aligning mergesort for a level 2 cache alone is
counter-productive - aligning it for level 1 and level 2 is significantly
faster; it is possible to avoid using a sentinel and still remove the bounds
check from a standard 2-heap, simply by reordering the steps involved;
sequential searches are far cheaper than binary searches across small lists, due
to a low branch misprediction rate. Bubblesort has terrible branch performance
in the middle, due to being partially sorted - it performs much worse than
selection sort as a result; multi-mergesort contains regular access patterns
which can be exploited by a 2-level adaptive branch predictor; in almost all
other cases, a bimodal predictor out-performs a 2-level adaptive predictor. A
median-of-3 quicksort outperforms other medians; radix sort has no branch
mispredictions; shellsort has easily calculable branch mispredictions, and
performs very similarly to an $O(NlogN)$ so.

In addition, we provide an analysis of most major sorts currently used, a
comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
sorts and of cache conscious sorts. We present branch prediction performance
figures for most major sorts, and adapt cache-conscious strategies from
mergesort to make an improved radixsort the fastest performing sort in our
tests.

>>>>

Among them is the fact that bimodal branch predictors have performed better than
two-level adaptive ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by two-level adaptive predictors - do
not exist, a bimodal predictors perform better.

@@

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a four
heap lowered our instruction count. Our level 2 misses also concurred with
LaMarca, with only one difference: memory-tuned heapsort, being an out-of-place sort,
fills the cache fully with a smaller number of keys than base heapsort, which is
in-place. LaMarca's results do not show this. Nor do they show a difference in
cache misses between these two sorts when the array does fit in the cache.
LaMarca's base heapsort has the characteristics of an out-of-place heapsort.
However, our results show that his implicit heap improvements have led to a
significantly faster heapsort.

@@

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $O(H_N-1)$ branch mispredictions per key, which
applies to long arrays, though short arrays become more predictable the longer
they get, as in heapsort; tiling mergesort can be done in two steps, reducing
the cache miss count; aligning mergesort for a level 2 cache alone is
counter-productive - aligning it for level 1 and level 2 is significantly
faster; it is possible to avoid using a sentinel and still remove the bounds
check from a standard 2-heap, simply by reordering the steps involved;
sequential searches are far cheaper than binary searches across small lists, due
to a low branch misprediction rate. Bubblesort has terrible branch performance
in the middle, due to being partially sorted - it performs much worse than
selection sort as a result; multi-mergesort contains regular access patterns
which can be exploited by a two-level adaptive branch predictor; in almost all
other cases, two-level branch predictors are measurably less accurate than
simple bimodal predictors at predicting branches in sorting algorithms. A
median-of-3 quicksort outperforms other medians; radix sort has no branch
mispredictions; shellsort has easily calculable branch mispredictions, and
performs very similarly to an $O(NlogN)$ so.

In addition, we provide an analysis of most major sorts currently used, a
comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
sorts and of cache conscious sorts. We present branch prediction performance
figures for most major sorts, and adapt cache-conscious strategies from
mergesort to make an improved radixsort the fastest performing sort in our
tests, and to make mergesort itself significantly faster.

### 2005-08-31 docs/tech_report/elementary.tex
lots of minor changes
david's final changes

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log(N)$\footnote{This seems like an error, since $H_N$ is very similar to
$log_eN$. It is not. We mean that $H_N$ is very similar to $log_2N$.}. Thus, in
the average case, we can expect that the comparison branch will mostly go in the
same direction, making the branch highly predictable.

@@

The number of misses for bubblesort and shakersort are so huge that we had to
plot them on a separate chart. Nearly one in four comparisons are misses. As the
data sets gets large, there are nearly 10,000 branch misses involved in sorting
a single key. The improved bubblesort had exactly the same number of misses and
branches. The improved shakersort has a lot less branches, but the same number
of misses. This is because skipping over sorted keys is predictable from the
first branch miss.

>>>>

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log(N)$. Thus, in the average case, we can expect that the comparison branch
will mostly go in the same direction, making the branch highly predictable.

@@

The number of misses for bubblesort and shakersort are so huge that we had to
plot them on a separate chart. Nearly one in four comparisons are misses. As the
data sets gets large, there are nearly 10000 branch misses involved in sorting
a single key. The improved bubblesort had exactly the same number of misses and
branches. The improved shakersort has a lot less branches, but the same number
of misses. This is because skipping over sorted keys is predictable from the
first branch miss.

### 2005-08-31 docs/tech_report/heap.tex
david's final changes

The relationship between this number and $H_N-1$ is relatively straightforward.
$H_N-1$ represents the cumulative number of changes in the minimum. Over a long
array, this approximates the number of branch mispredictions, as almost every
branch will not be taken, and the predictor will always predict not-taken.

Meanwhile, Figure \ref{memory-tuned heapsort curve} shows us the rate of correct
predictions, and each point is calculated using a separate branch predictor. As
a result, we expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the Cartesian
coordinates of each point on the graph. $1-y$ is the number of mispredictions
from the graph, while $\frac{1}{x+1}$ is the increase in $H_N$ for step $N$.

>>>>

The relationship between these numbers and $H_N-1$ is relatively straighforward.
$H_N-1$ represents the cumulative number of changes in the minimum. There first
element is guaranteed to be the minimum. The second element has a 50\% chance of
being smaller than the first. If we examine a third element, then the chances of
that element being smaller than two other elements is one in three. Therefore,
this branch will be taken one third of the time. As the number of elements we
examine increases, the probability of the nth element being smaller than all
elements seen so far is $\frac{1}{n}$.

Meanwhile, Figure \ref{memory-tuned heapsort curve} shows the rate of correct
predictions. The loop has been fully unrolled, so there is a separate branch
(with its own predictor) for each of the seven comparisons.  As a result, we
expect $1-y = \frac{1}{x+1}$, where $x$ and $y$ are the Cartesian coordinates of
each point on the graph. $1-y$ is the number of mispredictions from the graph,
while $\frac{1}{x+1}$ is the increase in $H_N$ for step $N$.

An interesting question is whether fully unrolling this loop, and thus
separating the comparison branch into seven separate branches, has an impact on
branch prediction. Essentially, there are two effects at work if we replace the
sequence of branches with a single branch. First, the single branch predictor
will spend most of its time in the state where it predicts not-taken, rather
than taken. This is because all branches apart from the first are biased in the
taken direction. This should improve accuracy. On the other hand, on those
occasions when the bi-modal branch predictor is pushed towards predicting taken,
the next branch is highly likely to be mispredicted. This is especially true
because the first two or three branches are most likely to push the predictor in
the taken direction, whereas the later branches are those that are mostly likely
to result in a misprediction in this state (because they are almost always
not-taken).

We investigated the trade-offs in these two effects and we discovered that using
a single branch does indeed cause the branches for the ealier comparisons to
become more predictable. However, this improvement is outweighed by a loss in
predictability of the later comparisons. The overall prediction accuracy fell by
around 2\% when we used a single branch, rather than unrolling the loop. This
result is interesting because it shows that relatively small changes in the way
that an algorithm is implemented can have an effect on branch prediction
accuracy. In this case, unrolling does not just remove loop overhead. It also
improves the predictability of the branches within the loop.
